Strengths and Weaknesses of Recurrent Quantification Analysis in the context of Human-Humanoid Interaction

### 2019-03-22 docs/manuscript/main.tex
[first draft]

\begin{abstract}
Movement variability is defined as the variations that occur in motor
performance across multiple repetitions of a task and such behaviour is an
inherent feature within and between each persons' movement. Quantifying
movement variability is still an open problem, particularly when traditional
methods in time domain and frequency domain fail to detect tiny modulations in
frequency or phase for time series. We therefore consider methodologies from
nonlinear dynamics such as reconstructed state space (RSS), uniform time-delay
embedding (UTDE), recurrence plots (RPs) and recurrence quantification analysis
(RQA) metrics. These algorithms require time series measured with sensors that
provide well sampled data with little noise. However, for this study, we are
interested in the analysis of data collected through wearable inertial sensors
(IMUs) and its post-processing effects in each of the nonlinear dynamics
methodologies. Twenty right-handed healthy participants (mean and standard
deviation (SD) age of mean=19.8 (SD=1.39)) performed an experiment in
human-humanoid imitation activities (H-H-I). Then IMUs were attached to the
participants and the humanoid robot performed simple vertical and horizontal
arm movements in normal and faster speed. With four window lengths and three
levels of smoothed time series, we found visual differences in the patterns of
RSSs and RPs. Then using metrics of RQA, we find out that the type of movements
and the level of smoothness affects those metrics. In particular, we found that
entropy values from RQA were well distributed and presenting variation in all
the conditions for time series. Hence, we demonstrated the potential of
nonlinear techniques to quantify human movement variability in the context of
H-H-I can enhance the development of better diagnostic tools for various
applications rehabilitation, sport science or for new forms of human-robot
interaction.
%280
\end{abstract}

@@

\subsection*{Recurrences Plots}
Considering the time series of Fig~\ref{fig:ts}, 
Recurrence Plots were computed for horizontal arm movements (Fig~\ref{fig:rp_aH}) and
for vertical arm movements (Fig~\ref{fig:rp_aV}) 
using the average embedding parameters ($m=6$, $\tau=8$) 
and an recurrence threshold of $\epsilon=1$.
With regard to the selection of recurrence threshold,
Marwan et al. \cite{marwan2011} pointed out that choosing an appropriate 
recurrence threshold is crucial to get meaningful representations in RPs, 
however, for this work where quantifying movement variability is our aim,
we do not compute recurrence thresholds as as long as it is able to 
represent the dynamical transitions in each of the time series.

As similar as with the Reconstructed State Spaces, 
the differences in the RPs can be easily noticed by eye 
for different time series (Figs~\ref{fig:rp_aV} and ~\ref{fig:rp_aH}), 
which lead us to apply Recurrence Quantification Analysis 
to have an objective quantification of each of the time series.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rps/pdf/rp_aH}
\caption{
	{\bf RPs for horizontal arm movements.}	
	Recurrence plots were computed with 
	embedding parameters $m=6$, $\tau=8$ and $\epsilon=1$.
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:rp_aH}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rps/pdf/rp_aV}
\caption{
	{\bf RPs for vertical arm movements.}	
	Recurrence plots were computed with 
	embedding parameters $m=6$, $\tau=8$ and $\epsilon=1$.
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:rp_aV}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsection*{Recurrence Quantification Analysis} \label{ch6:rqas}
Considering the RPs for 20 participants performing four activities 
(HN, HF, VN and VF) with sensors attached to the human (HS01) and to the 
humanoid robot (RS01) and with the increase of smoothness of time series 
(sg0zmuvGyroZ, sg1zmuvGyroZ and sg2zmuvGyroZ), 
we hence compute four metrics of RQA metrics (REC, DET, RATIO and ENTR) with 
embedding parameters $m=6$, $\tau=8$ and recurrence threshold $\epsilon=1$.

%\newpage
\subsubsection*{REC values}
It can be seen in the box plots of Figs~\ref{fig:RQABP}(A) that REC values, 
representing the \% of black dots in the RPs, 
are more spread for HN and VN movements (higher interquartile range) 
than HF and VF movements (lower interquartile range) for HS01 sensor. 
In contrast, REC values for RS01 sensor present little variation 
(interquartile range of 0.01).
With regard to the increase of smoothness of time series 
(sg0, sg1 and sg2), REC values present little 
variation as the smoothness is increasing for time series from HS01 
(changes of mean values (rhombus)) while REC values are more affected with 
the smoothness for data from RS01 
(see the incremental changes of mean values (rhombus)).
%See Figs~\ref{fig:rec_aH} and \ref{fig:rec_aV} in appendix \ref{appendix:e:ep} 
%for more details about individual REC values for each participant.

\subsubsection*{DET values}
Figs \ref{fig:RQABP}(B) illustrate DET values, 
representing predictability and organisation of the RPs, 
which change very little (interquartile range is around 0.1) 
for type of movement, level of smoothness or type of sensor.
%See Figs~\ref{fig:det_aH} and \ref{fig:det_aV} in appendix \ref{appendix:e:ep} 
%for more details about individual DET values for each participant.

\subsubsection*{RATIO values}
Figs \ref{fig:RQABP}(C) present RATIO values, representing dynamic transitions, 
for horizontal and vertical movements.
It can be seen that RATIO values for HS01 sensor vary less 
for HN movements (interquartile range around 2)
than HF movements (interquartile range around 5).
%which is a similar behaviour of RATIO values for oRS01 sensor.
It can also be noticed a decrease of variation in RATIO values as the 
smoothness of the time series is increasing (grey rhombus).
%See Figs~\ref{fig:ratio_aH} and \ref{fig:ratio_aV} in appendix 
%\ref{appendix:e:ep} 
%for more details about individual RATIO values for each participant.

\subsubsection*{ENTR values}
Figs \ref{fig:RQABP}(D) show ENTR values, representing the complexity of 
the structure the time series, for both horizontal and vertical movements.
ENTR values for HS01 sensor show more variation 
(interquartile range around 0.5)
than ENTR values for RS01 sensor which appear 
to be more constant (interquartile range 0.1).
It can also be said that the smoothness of time series affects
each of the axis by an increase of mean values (see gray rhombos).
%See Figs~\ref{fig:entr_aH} and \ref{fig:entr_aV} in appendix
%\ref{appendix:e:ep}
%for more details about individual ENTR values for each participant.

@@

%
%
%\subsection*{RQA metrics with different embedding parameters, 
%	recurrence thresholds, window lengths, levels of smoothness, 
%	and time series structures.}
%Zbilut et al. \cite{zbilut1992} established RQA metrics with the aim of 
%determining embedding parameters, their method consisted on creating 3D 
%surfaces with RQA metrics with an increase of embedding parameters 
%($m$ and $\tau$), then Zbilut et al. \cite{zbilut1992} explored 
%fluctuations and gradual changes in the 3D surfaces that provide information 
%about the embeddings. Much recently, Marwan et al. \cite{marwan2015} 
%created 3D surfaces for visual selection of not only embedding parameters 
%but also recurrence thresholds. Following same methodologies, we explored 
%the stability and robustness of RQA metrics (REC, DET, RATIO and ENTR)
%using 3D surfaces by an unitary increase of the pair embedding 
%parameters ($0 \ge m \le 10$, $0 \ge \tau \le 10$) and a decimal increase 
%of 0.1 for recurrence thresholds ($ 0.2 \ge \epsilon \le 3 $) 
%(Fig.~\ref{fig:topo_rqas}). We also computed 3D surfaces of RQA metrics 
%for different sensors and different activities 
%(Fig.~\ref{fig:topo_sensoractivities}). RQA metrics are also affected by 
%the window length where for example four window lengths of 100, 250, 500 
%and 750 samples (Fig.~\ref{fig:topo_sensoractivities}). Three level of 
%smoothness were computed for RQA metrics showing smoothed 3D surfaces and 
%the level of smoothness increase (Fig.~\ref{fig:topo_smoothness}).
%Similarly, 3D surfaces of RQA metrics were also computed for three 
%participants (Fig.~\ref{fig:topo_participants}).
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/rqas}
%%\includegraphics[width=1.0\textwidth]{rqas}
%    \caption{
%	{\bf 3D surfaces for RQA metrics.}
%	3D surfaces for REC, DET, RATIO and ENTR values with increasing 
%	pair embedding parameters ($0 \ge m \le 10$, $0 \ge \tau \le 10$) 
%	and recurrence thresholds (  $ 0.2 \ge \epsilon \le 3 $).
%	RQA metrics values for time series of participant p01 using HS01 sensor, 
%	HN activity and sg0zmuvGyroZ axis and 500 samples window length.
%        R code to reproduce the figure is available from \cite{hwum2018}.
%	}
%\label{fig:topo_rqas}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/sa}
%%\includegraphics[width=1.0\textwidth]{sa}
%    \caption{
%	{\bf 3D surfaces of RQA metrics for sensors and activities.}
%	3D surfaces with increasing embedding parameters and recurrence 
%	thresholds are for HS01 and RS01 sensors of HN, HF, VN and VF 
%	activities. RQA metrics values are for time series of 
%	participant p01 for sensors (HS01 and RS01), 
%	activities (HN, HF, VN and VF) and for 	sg0zmuvGyroZ axis with 
%	500 samples window length. 
%	R code to reproduce the figure is available from \cite{hwum2018}.
%       }
%\label{fig:topo_sensoractivities}
%\end{figure}
%%%---------------------------------(FIGURE)-------------------------------------
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/w}
%%\includegraphics[width=1.0\textwidth]{w}
%    \caption{
%	{\bf 3D surfaces for RQAs metrics with four window lengths.}
%	3D surfaces of RQAs metric values with increasing embedding 
%	parameters and recurrence thresholds are for four window lengths 
%	(w100, w250, w500 and  w750).
%	RQA metrics values are for time series of participant p01 
%	using HS01 sensor, HN activity and sg0zmuvGyroZ axis.
%	R code to reproduce the figure is available from \cite{hwum2018}.
%        }
%\label{fig:topo_sensoractivities}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/s}
%%\includegraphics[width=1.0\textwidth]{s}
%    \caption{
%	{\bf 3D surfaces for RQA metrics with three levels of smoothness.}
%       	3D surfaces of RQAs metric values with increasing embedding 
%	parameters and recurrence thresholds are for three levels of 
%	smoothness (sg0zmuvGyroZ, sg1zmuvGyroZ and sg1zmuvGyroZ).
%	RQA metrics values are for time series of participant p01 using 
%	HS01 sensor, HN activity and 500 samples window length.
%	R code to reproduce the figure is available from \cite{hwum2018}.
% }
%\label{fig:topo_smoothness}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------
%
%
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/p}
%%\includegraphics[width=1.0\textwidth]{p}
%    \caption{
%	{\bf 3D surfaces for RQA metrics with three participants.}
%       	3D surfaces of RQAs metric values for participants p01, p02 and p03 
%	with increasing embedding parameters and recurrence thresholds.
%	RQA metrics values are for time series of HS01 sensor, 
%	HN activity and 500 samples window length.
%	R code to reproduce the figure is available from \cite{hwum2018}.
% }
%\label{fig:topo_participants}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------
%
%
%
%
%
%
%
%
%
%
%
%%*******************************************************************************
%%*******************************************************************************
%%*******************************************************************************
%\section*{Discussion}
%%The Discussion should be succinct and must not contain subheadings.
%
%It is evidently that time series from different sources (participants, 
%movements, axis type, window length or levels of smoothness) 
%presents visual differences for embedding parameters and therefore for RRSs. 
%For which, the selection of embedding parameters was our first challenge 
%where we computed embedding parameters for each time series and then computed 
%a sample mean over all time series in order to get two embedding parameters 
%to compute all RSSs with its corresponded type of movement. Then we found 
%that the quantification of variability with regard to the shape of the 
%trajectories in RSSs requires more investigation since our original 
%proposed method base on euclidean metric failed to quantify those 
%trajectories. Specially, for trajectories which were not well unfolded. 
%With that in mind, we proceed to take advantage of four RQA metrics 
%(REC, DET, RATIO and ENTR) in order to avoid any subjective interpretations 
%or personal bias with regard to the evolution of the trajectories in RSSs.
%
%
%\subsection*{RQA metrics with fixed parameters}
%Considering that RQA metrics were computed with fixed embedding parameters 
%($m=6$ and $\tau=8$) and recurrence thresholds ($\epsilon=1$), we found 
%the following. REC values, which represents the \% of black points in the RPs, 
%were more affected with and increase in normal speed movements (HN and VN) 
%than faster movements (HF and VF) for the sensor attached to the participants 
%(HS01). Such decrease of REC values from normal speed to faster speed 
%movements is also presented in data from sensor attached to the robot (RS01), 
%and little can be said with regard to the dynamics of the time series coming 
%from RS01. Similarly, DET values, representing predictability and 
%organisation in the RPs, present little variation in the any of the time 
%series where little can be said. In contrast, RATIO values, which represent 
%dynamic transitions, were more variable for faster movements (HF and VF) 
%than normal speed movements (HN and VN) with sensors attached to the 
%participants (HS01). For data coming from sensors attached to the robot 
%(RS01), RATIO values from horizontal movements (HN, HF) appear to vary 
%more than values coming from vertical movmentes (VN, VF).
%With that, it can be said that RATIO values can represent a bit better
%than REC or DET metrics for the variability of imitation activities in 
%each of the conditions for time series.
%In the same way, ENTR values for HN were higher than values for HF
%and ENTR values varied more for sensor attached to participants 
%than ENTR values for sensors of the robot. It is evidently that 
%the higher the entropy the more complex the dynamics are, 
%however, ENTR values for HN appear a bit higher than HF values, 
%for which we believe this happens because of the structure the time series
%which appear more complex for HN than  HF movements which presented a 
%more consistence repetition.
%
%We observed that some RQA metrics are affected by the smoothness of data.
%For which, we also explored the effect of smoothness of raw-normalised data 
%where, for example, REC and DET values were not affected by the smoothness 
%of data since these seemed to be constants. However, for RATIO values, 
%the effect of smoothness can be noticed with a slight decrease of amplitude 
%in any of the time series conditions which is also presented with ENTR values.
%
%
%\subsection*{RQA metrics with different parameters}
%Patterns in RPs and metrics for RQA are independent of embedding dimension 
%parameters \cite{iwanski1998}, however, that is not the case when using 
%different recurrence thresholds. Such changes of recurrence threshold values 
%can modify the patterns in RPs and therefore the values of RQA metrics.
%We therefore computed 3D surfaces to explore the sensibility and robustness of 
%embedding parameters and recurrence threshold in RQA  metrics. Following the 
%same methodology of computing 3D surfaces, we also considered variation of 
%window length size to present RQA metrics dependencies with embedding 
%parameters, recurrence thresholds and window length size.
%
%
%
%\section*{Conclusions}
%We conclude that using a different level of smoothness for time series help us 
%to visualise and to quantify the variation of movements between participants 
%using RSSs, RPs and RQA. Also, it is important to mention that some RQA's 
%metrics (e.g. DET and  ENTR) are more robust to the effect of smoothness 
%of time series.
%
%Althought our our experiment is limited to twenty healthy right-handed 
%participants of a range age of mean 19.8 SD=1.39, RQA metrics can quantify 
%human movement variability. With that in mind, we conclude that 
%quantification of human-humanoid imitation activities is possible for 
%participants of different ages, state of health and anthropomorphic features.
%
%
%%
%%Such understanding and measurement of movement variability using
%%cheap wearable inertial sensors lead us to have a more intuitive selection of parameters
%%to reconstruct the state spaces and to create meaningful interpretations
%%of the recurrence plots and the results of the metrics with recurrence quantification 
%%analysis. 
%%
%
%In general, activity type, window length and structure of the time series 
%affects the values of the metrics of RQA for which certain RQA metrics 
%are better to describe determined type of movement.
%Using determined RQA metrics depends on what one want to quantify, for instance, 
%one can find predicability, organisation of the RPs, dynamics transitions, 
%or complexity and determinism.
%
%
%%
%%However, we believe that further investigation is required to find the 
%%right balance between the level of smoothness of the signal
%%and defining what is the aim of RQA where 
%%its representations using RSS, RP and RQA.
%%Specially, where the level of smoothness does not affect 
%%the variation of each of the movements quantification. 
%%%We can also conclude that finding the right balance between 
%%%smoothness and the raw data to capture movement variability is 
%%%a still a problem that has many avenues for exploration.
%%
%
%Similarly, such differences in time series created differences in each of the 
%RQA metrics, for instance, RATIO and ENTR are helpful to distinguish 
%differences in any of the categories of the time series (sensor, activity, 
%level of smoothness and number of participant), however for certain time series 
%(data from the sensor attached to the robot) seemed to have little variations 
%between each of the participants. The latter phenomena was in a way evidently
%as robot degrees of freedom did not allow it to move with a wide range of 
%variability. 
%
%
%

>>>>

\begin{abstract}
Human movement variability occurs in motor performance 
across multiple repetitions of a task and such behaviour is an
inherent feature within and between each persons' movement. Quantifying
movement variability is still an open problem, particularly when traditional
methods in time domain and frequency domain fail to detect tiny modulations in
frequency or phase for time series. 
For this work, we hence investigate methodologies from nonlinear dynamics such as 
reconstructed state space (RSS), uniform time-delay embedding (UTDE), 
recurrence plots (RPs) and recurrence quantification analysis (RQA) metrics. 
Particularly, we are interested in the weaknesses and robustness 
of nonlinear dynamics tools from raw and post-processed data of 
wearable inertial sensors (IMUs). 
In the reported experiment, twenty right-handed healthy participants 
imitated simple vertical and horizontal arm movements in normal 
and faster speed from an humanoid robot.
With four window lengths and three levels of smoothed time series, 
we found visual differences in the patterns of RSSs and RPs. 
Then using metrics of RQA, we find out that the type of movements
and the level of smoothness affects those metrics. 
Our contribution for this work is that entropy values from RQA were 
well distributed and presenting variation in all the conditions for time series. 
This work have then the potential to enhance the development of 
better diagnostic tools for various applications rehabilitation, 
sport science or for new forms of human-robot interaction.
\end{abstract}
%226

@@

\subsection*{Recurrences Plots}
Recurrence Plots (RP) were computed for horizontal arm movements (Fig~\ref{fig:rp_aH}) and 
vertical arm movements (Fig~\ref{fig:rp_aV}) 
using the average embedding parameters ($m=6$, $\tau=8$) 
and an recurrence threshold of $\epsilon=1$.

%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rps/pdf/rp_aH}
\caption{
	{\bf RPs for horizontal arm movements.}	
	Recurrence plots were computed with 
	embedding parameters $m=6$, $\tau=8$ and $\epsilon=1$.
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:rp_aH}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rps/pdf/rp_aV}
\caption{
	{\bf RPs for vertical arm movements.}	
	Recurrence plots were computed with 
	embedding parameters $m=6$, $\tau=8$ and $\epsilon=1$.
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:rp_aV}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsection*{Recurrence Quantification Analysis} \label{ch6:rqas}
%Considering the RPs for 20 participants performing four activities 
%(HN, HF, VN and VF) with sensors attached to the human and 
%to the humanoid robot (HS01, RS01) and the increase of smoothness 
%of time series (sg0zmuvGyroZ, sg1zmuvGyroZ and sg2zmuvGyroZ),
Similarly to RP, we computed four RQA metrics (REC, DET, RATIO and ENTR) 
with embedding parameters $m=6$, $\tau=8$ and 
recurrence threshold $\epsilon=1$.

\subsubsection*{REC values}
It can be seen in the box plots of Figs~\ref{fig:RQABP}(A) that REC values, 
representing the \% of black dots in the RPs, 
are more spread for HN and VN movements (higher interquartile range) 
than HF and VF movements (lower interquartile range) for HS01 sensor. 
In contrast, REC values for RS01 sensor present little variation 
(interquartile range of 0.01).
With regard to the increase of smoothness of time series 
(sg0, sg1 and sg2), REC values present little 
variation as the smoothness is increasing for time series from HS01 
(changes of mean values (rhombus)) while REC values are more affected with 
the smoothness for data from RS01 
(see the incremental changes of mean values (rhombus)).
%See Figs~\ref{fig:rec_aH} and \ref{fig:rec_aV} in appendix \ref{appendix:e:ep} 
%for more details about individual REC values for each participant.

\subsubsection*{DET values}
Figs \ref{fig:RQABP}(B) illustrate DET values, 
representing predictability and organisation of the RPs, 
which change very little (interquartile range is around 0.1) 
for type of movement, level of smoothness or type of sensor.
There is also increase of DET values as the smoothness of the signal increase 
(see the incremental changes of mean values (rhombus)).
It can be noted that the interquartile range for faster movements
(HF and VF) with no smoothing (sg0) is lower than the other
levels of smoothness (sg1 and sg2).
%See Figs~\ref{fig:det_aH} and \ref{fig:det_aV} in appendix \ref{appendix:e:ep} 
%for more details about individual DET values for each participant.

\subsubsection*{RATIO values}
Figs \ref{fig:RQABP}(C) present RATIO values, representing dynamic transitions, 
for horizontal and vertical movements.
It can be seen that RATIO values for HS01 sensor vary less 
for HN movements (interquartile range around 2)
than HF movements (interquartile range around 5).
%which is a similar behaviour of RATIO values for oRS01 sensor.
For faster movements (HF and VF), it can be noted a decrease of 
RATIO values as the smoothness of the time series is increasing (rhombus).
For normal movements (HN and VN), the decrease of RATIO values
is less evident than faster movements (see rhombus).
%See Figs~\ref{fig:ratio_aH} and \ref{fig:ratio_aV} in appendix 
%\ref{appendix:e:ep} 
%for more details about individual RATIO values for each participant.

\subsubsection*{ENTR values}
Figs \ref{fig:RQABP}(D) show ENTR values, Shannon entropy values 
which represent the complexity of 
the structure the time series, 
for type of movement, level of smoothness or type of sensor.
ENTR values for HS01 sensor show a higher variation  
(interquartile range around 0.5)
than ENTR values for RS01 sensor which appear 
to be more constant (interquartile range 0.1).
It can also be noted the increase of smoothness of time series 
also is presented with an increase of ENTR values (see gray rhombus).
%See Figs~\ref{fig:entr_aH} and \ref{fig:entr_aV} in appendix
%\ref{appendix:e:ep}
%for more details about individual ENTR values for each participant.

@@

\subsection*{3D RQA ENTR}
Choosing an appropriate recurrence threshold is crucial to get 
meaningful representations in RP and RQA, however, as previously shown, 
one can selected the default recurrence threshold of $\epsilon=1$, 
as long as it is able to represent the dynamical transitions 
in each of the time series \cite{marwan2011}. %%Marwan et al. \cite{marwan2011} 
However, to show how that the selection of recurrence threshold affects RQA
values, 
we computed ENTR values of RQA metrics for different embedding parameters
$[1,10]= \{ m \in \mathbb{R} | 1 \le m \le 10  \} $,
$[1,10]= \{ \tau \in \mathbb{R} | 1 \le \tau \le 10  \} $
with an increase of 1
and different recurrence thresholds 
$[0.2,3]= \{ \epsilon \in \mathbb{R} | 0.2 \le \epsilon \le 3.0  \} $
with increments of 0.1.
For easily visualisation, Figs \ref{fig:RQA-IND} only show 3D RQA for 
three recurrence thresholds $\epsilon=1, 2, 3$ with three levels of 
smoothness (sg0, sg1, sg2). It can be noted in Figs \ref{fig:RQA-IND} 
that the increase of recurrence threshold is associated to the increase 
of ENTR values in any of the levels of smoothness (see values of the ENTR bar).
One can also see that the maximum values of ENTR 
for $\epsilon=1$ and sg0 are for embedding dimensions of 2 
and then as $\epsilon$ increases the maximum values of ENTR are for
an associated increase of in the embedding dimension.
It can also be noted that the increase of level of smoothness (from sg0 to sg2) 
is associated with both the increase of ENTR values and its smoothed 3D shape.

%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/rqa/pdf/rqa_epsilons}
    \caption
	[3D RQA ENTR values]{
	{\bf 3D RQA ENTR values.}
	RQA ENTR values 
	for different embedding parameters
	$[1,10]= \{ m \in \mathbb{R} | 0 \le m \le 10  \} $,
	$[1,10]= \{ \tau \in \mathbb{R} | 0 \le \tau \le 10  \} $
	with an increase of 1
	and different recurrence thresholds $\epsilon=1, 2, 3$.
	RQA ENTR values are for $p03$, sensor HS01, of a window size of 10-secs (500 samples).
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:RQA-IND}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

%of ENTR and plotted 3D surfaces using an unitary 
%and a decimal increase of 0.1 for recurrence thresholds 
% (Fig.~\ref{fig:topo_rqas}). 
%Hence, Fig.~\ref{fig:topo_rqas}(A) shows an increase for 
%REC values, the percentage of black docs in the RP, 
%as the recurrence threshold increase,
%while the variation for embedding parameters creates little decrease 
%of REC values as the embedding dimensions.
% increase and even a more slighter 
%decrements of REC values for the increase of $\tau$.
%For the 3D surface of DET values (Fig.~\ref{fig:topo_rqas}(B)), 
%representing 
%predictability and organisation of the RPs, one can be noted a plateau
%for DET values near to 1 for embedding dimension parameters of less 
%than 5 and recurrence threshold values of greater than 2 (red surface). 
%It can also be observed that the increases of delay embedding made 
%the DET values increase so as to make an cascade effect in the surface 
%along with the increase of dimension embedding $m$.
%For RATIO values, representing dynamic transitions, 
%Fig.~\ref{fig:topo_rqas} shows that the 3D surface present 
%a plateau (blue surface) of RATIO values 
%near to zero for recurrence thresholds greater than 1.0, while 
%fluctuations are more evident for recurrence thresholds of less than 1.0,
%particularly it can also be noted an increase in the fluctuations of 
%RATIO values as the embedding dimension is increasing. 
%For ENTR values in Fig.~\ref{fig:topo_rqas}(D), 
%representing the complexity of the structures in time series, 
%one can be noted that the increase of recurrence threshold is, 
%not strictly proportional to the increase of ENTR values. 
%It can also be observed in Fig.~\ref{fig:topo_rqas}(D)
%that the increase of delay embeddings affects 
%little the ENTR values for embedding dimensions of 1, while 
%for higher values of embedding dimensions 
%there is a decrease of ENTR values, and 
%there is a decrease of ENTR values as delay dimension value is increasing.
%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}
%\centering
%\includegraphics[width=1.0\textwidth]{rqas}
%    \caption
%	[3D surfaces for RQA metrics]{
%	{\bf 3D surfaces for RQA metrics.}
%	3D surfaces for (A) REC, (B) DET, (C) RATIO and (D) ENTR values 
%	with increasing pair of embedding parameters 
%	($0 \ge m \le 10$, $0 \ge \tau \le 10$) 
%	and recurrence thresholds (  $ 0.2 \ge \epsilon \le 3 $).
%	RQA metrics are computed with the time series of participant $p01$ using 
%	HS01 sensor, HN activity, sg0zmuvGyroZ axis and 500 samples 
%	for window size length.
%	Code and data to reproduce the figure is available from \cite{srep2019}.
%	}
%\label{fig:topo_rqas}
%\end{figure}
%%---------------------------------(FIGURE)------------------------------------
%

\subsection*{Other effects on 3D RQA ENTR values}
% metrics with different embedding parameters, 
%	recurrence thresholds, window lengths, levels of smoothness, 
%	and time series structures.}

%Zbilut et al. \cite{zbilut1992} established RQA metrics with the aim of 
%determining embedding parameters, their method consisted on creating 3D 
%surfaces with RQA metrics with an increase of embedding parameters 
%($m$ and $\tau$), then Zbilut et al. \cite{zbilut1992} explored 
%fluctuations and gradual changes in the 3D surfaces that provide information 
%about the embeddings. Much recently, Marwan et al. \cite{marwan2015} 
%created 3D surfaces for visual selection of not only embedding parameters 
%but also recurrence thresholds. Following same methodologies, we explored 
%the stability and robustness of RQA metrics (REC, DET, RATIO and ENTR)
%using 3D surfaces by an unitary increase of the pair embedding 
%parameters ($0 \ge m \le 10$, $0 \ge \tau \le 10$) and a decimal increase 
%of 0.1 for recurrence thresholds ($ 0.2 \ge \epsilon \le 3 $) 
%(Fig.~\ref{fig:topo_rqas}). 
We also computed 3D surfaces of RQA metrics for different sensors 
and different activities (Fig~\ref{fig:3dRQAENTR_sensoractivities}).
From the maximum values of ENTR (lateral bars), one can see 
a decrease of ENTR values for activities going from normal to 
faster velocity and from human sensor (HS01) to robot sensor (RS01) 
(See Fig~\ref{fig:3dRQAENTR_sensoractivities}). 
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rqa/pdf/rqa_sensors_activities}
    \caption{
	{\bf 3D surfaces of RQA ENTR metrics for sensors and activities.}
	RQA ENTR values with embedding parameters
	$[1,10]= \{ m \in \mathbb{R} | 0 \le m \le 10  \}$,
	$[1,10]= \{ \tau \in \mathbb{R} | 0 \le \tau \le 10  \}$
	for different activities and sensors. 
	RQA ENTR values are for $p03$, sg0 and window size of 10-secs (500 samples).
	Code and data to reproduce the figure is available from \cite{srep2019}.
       }
\label{fig:3dRQAENTR_sensoractivities}
\end{figure}
%%---------------------------------(FIGURE)-------------------------------------

%RQA metrics are also affected by 
%the window length where for example four window lengths of 100, 250, 500 
%and 750 samples (Fig.~\ref{fig:topo_sensoractivities}).

%Similarly, 3D surfaces of RQA metrics were also computed for three 
%participants (Fig.~\ref{fig:topo_participants}).

%X Three level of 
%X smoothness were computed for RQA metrics showing smoothed 3D surfaces and 
%X the level of smoothness increase (Fig.~\ref{fig:topo_smoothness}).

%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/rqas}
%%\includegraphics[width=1.0\textwidth]{rqas}
%    \caption{
%	{\bf 3D surfaces for RQA metrics.}
%	3D surfaces for REC, DET, RATIO and ENTR values with increasing 
%	pair embedding parameters ($0 \ge m \le 10$, $0 \ge \tau \le 10$) 
%	and recurrence thresholds (  $ 0.2 \ge \epsilon \le 3 $).
%	RQA metrics values for time series of participant p01 using HS01 sensor, 
%	HN activity and sg0zmuvGyroZ axis and 500 samples window length.
%        R code to reproduce the figure is available from \cite{hwum2018}.
%	}
%\label{fig:topo_rqas}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------

%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/w}
%%\includegraphics[width=1.0\textwidth]{w}
%    \caption{
%	{\bf 3D surfaces for RQAs metrics with four window lengths.}
%	3D surfaces of RQAs metric values with increasing embedding 
%	parameters and recurrence thresholds are for four window lengths 
%	(w100, w250, w500 and  w750).
%	RQA metrics values are for time series of participant p01 
%	using HS01 sensor, HN activity and sg0zmuvGyroZ axis.
%	R code to reproduce the figure is available from \cite{hwum2018}.
%        }
%\label{fig:topo_sensoractivities}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/s}
%%\includegraphics[width=1.0\textwidth]{s}
%    \caption{
%	{\bf 3D surfaces for RQA metrics with three levels of smoothness.}
%       	3D surfaces of RQAs metric values with increasing embedding 
%	parameters and recurrence thresholds are for three levels of 
%	smoothness (sg0zmuvGyroZ, sg1zmuvGyroZ and sg1zmuvGyroZ).
%	RQA metrics values are for time series of participant p01 using 
%	HS01 sensor, HN activity and 500 samples window length.
%	R code to reproduce the figure is available from \cite{hwum2018}.
% }
%\label{fig:topo_smoothness}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------
%
%
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/p}
%%\includegraphics[width=1.0\textwidth]{p}
%    \caption{
%	{\bf 3D surfaces for RQA metrics with three participants.}
%       	3D surfaces of RQAs metric values for participants p01, p02 and p03 
%	with increasing embedding parameters and recurrence thresholds.
%	RQA metrics values are for time series of HS01 sensor, 
%	HN activity and 500 samples window length.
%	R code to reproduce the figure is available from \cite{hwum2018}.
% }
%\label{fig:topo_participants}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------
%
%
%
%
%
%
%
%
%
%
%
%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\section*{Discussion}
%The Discussion should be succinct and must not contain subheadings.
